# Barycenter Discriminant Analysis

Barycenter Discriminant Analysis is a robust version of Discriminant Analysis that is used like discriminant analysis - when multiple measurements describe a set of observations in which each observation belongs to one category from a set of priori defined categories. The goal of BADA is to combine the measurements to create new variables that best seperate the categories. These discriminant variables are also used to assign the original observations or new observations to the a-priori defined categories[@BADA]. 

The first step of BADA is to compute the barycenter of each of the N categories describing the rows. The $N$ by $J$ matrix of barycenters is computed as: $$\mathbf{R}= diag\{\mathbf{ Y^TM1\}}^{-1}\mathbf{Y^T MX}$$ where $Y$ is the design matrix for the categories describing the rows of $X$

The matrix $R$ is then analyzed using a GPCA under the constrints provided by the matrices $B$ and $W$. $$\mathbf{R=P \Delta Q^T}$$ with $\mathbf{P^T BP = Q^T WQ = I}$. The $N$ by $L$ matrix of factor scores for the categories is obtained as: $$\mathbf{F=P \Delta = RWQ}$$

The loadings describe the variables of the barycentric data matrix and are used to identify the variables important for the separation between the groups. 

For this analysis we use the `World Health` dataset and the heatmap of designmatrix * data is shown in figure \@ref(fig:BADAHeat)

```{r include=FALSE}
# Clean Start
rm(list = ls()) 
graphics.off() 

library(TInPosition)
require(ggplot2)

# Loading Dataset
WorldHealth <- read.csv("Data/WorldHealth.csv", row.names = 1)

# Cleaning Data 
names(WorldHealth) <- c("Fires", "Drownings", "Poisonings", "Falls",
                        "T.Chole_F", "T.Chole_M",
                        "BP_F", "BP_M",
                        "BMI_F","BMI_M",
                        "GenGovExp", "TotalExp_GDP", 
                        "PerCapitaExp_XchangeRate","PerCapita_PPP",
                        "IMM_MCV", "IMM_DTP3")

# Creating WorldHealth_Risk sub-table
WorldHealth_Risk <- WorldHealth[,c(5:10)]
designMatrix <- read.csv("Data/design_new.csv")

WorldHealth_Spending <- WorldHealth[,c(11:14)]


WorldHealth_Immunization <- WorldHealth[,c(15,16)]

# A function for plotting multiple plots 
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)

  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)

  numPlots = length(plots)

  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                    ncol = cols, nrow = ceiling(numPlots/cols))
  }

 if (numPlots==1) {
    print(plots[[1]])

  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))

    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))

      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}
```

```{r include=FALSE}
#  tepBADA.inference.battery() is a function from TInPosition package for doing BADA

resBADA <- tepBADA.inference.battery(WorldHealth_Risk, 
                                     scale = 'SS1',
                                     DESIGN = designMatrix$Region, 
                                     make_design_nominal = T, 
                                     center = T, 
                                     test.iters = 100, graphs = F)
```

```{r BADAHeat, fig.width=6, fig.height=6, out.width="50%", fig.align="center", fig.cap="BADA Heatmap", fig.pos="H", echo=FALSE, message=FALSE, warning=FALSE}
# RColorBrewer selects color palette 

require(RColorBrewer)
heatmap(resBADA$Fixed.Data$TExPosition.Data$X, 
        Rowv = NA, 
        Colv = NA, 
        col= colorRampPalette(brewer.pal(5, "Blues"))(256), 
        margins = c(10, 10))
```

```{r plotScreeBADA, include=FALSE}

# Create a function to plot the scree
# ev: the eigen values to plot. no default
# max.ev the max eigen value
#        needed because ExPosition does not return all ev
#        but only the requested one. but return all tau
#        so if max.ev is specified, it is used to recompute
#        all eigenvalues
# p.ep: the probabilities associated to the ev
# alpha: threshold for significance. Default = .05
# col.ns  = color for ns ev. Default is Green
# col.sig = color for significant ev. Default is Violet
PlotScree <- function(ev,p.ev=NULL,max.vp=NULL,
                      alpha=.05,
                      col.ns = '#006D2C',col.sig='#54278F',
                      title = "Explained Variance per Dimension"
){
  # percentage of inertia
  val.tau = (100*ev/sum(ev))
  Top.y = ceiling(max(val.tau)*.1)*10
  # if ev is already a percentage convert it back
  if (!is.null(max.vp)){ev = ev*(max.vp/ev[1])}
  p <- ggplot(ev, aes(x = c(1:5), y= ev))
  p + 
    geom_line() + ylab("Inertia Extracted by the Components") +
    scale_x_continuous(name='Dimensions', breaks = c(1,2,3,4,5))  +
    geom_line() + 
    scale_y_continuous(sec.axis = sec_axis(~./sum(ev)*100,name="Percentage of Explained Variance")) +
    geom_point(col=ifelse(p.ev<alpha, 'blue', 'indianred3'), size=2) +theme_get()
   
} # end of function PlotScree

```

## Eigenvalues/Variance 

The Scree plot for BADA is shown in figure \@ref(fig:BADAScree). From the figure we can see that BADA generated five components and out of them only two of them are important. On inference analysis none of them were found to be significant. 

```{r BADAScree, out.width="60%", fig.align="center", fig.cap="Scree Plot for BADA", fig.pos='H', fig.width=7.5, fig.height=4.5, echo=FALSE }
PlotScree(as.data.frame(resBADA$Fixed.Data$TExPosition.Data$eigs), p.ev = resBADA$Inference.Data$components$p.vals)
```

## Factor scores

IN BADA we can plot two factor map for our observations - one for individuals and other for groups (based on their mean). The factor map for groups is shown in figure \@ref(fig:BADAgroupmap) 

```{r BADAgroupmap, out.width="75%", fig.align="center", fig.pos="H", warning=FALSE, message=FALSE, fig.cap="BADA Factor Scores (Group)", echo=FALSE, fig.height=5, fig.width=8}
require(ggrepel)
FactorScore <- as.data.frame(resBADA$Fixed.Data$TExPosition.Data$fi[,c(1,2)])
FactorScore$labels <- labels(resBADA$Fixed.Data$TExPosition.Data$fi[,2])

p <- ggplot()

p + 
  # Plotting the points 
  geom_point(aes(x=FactorScore$V1,
                   y=FactorScore$V2)) +
  
  # Drawing confidence ellipse 
  stat_ellipse()+
  
  # Printing labels 
  geom_text_repel(aes(x=FactorScore$V1,
                      y=FactorScore$V2, 
                      label = FactorScore$labels))+
  
  # Scaling axis 
  scale_x_continuous(name = paste0("Component 1 Inertia: ",
                                            round(resBADA$Fixed.Data$TExPosition.Data$t[1],3), "%", 
                                   ", p=", resBADA$Inference.Data$components$p.vals[1]), 
                     limits = c(resBADA$Fixed.Data$Plotting.Data$constraints$minx, resBADA$Fixed.Data$Plotting.Data$constraints$maxx) ) +
  
  scale_color_discrete(name= "Regions")+
  
  scale_y_continuous(name = paste0("Component 2 Inertia: ", 
                            round(resBADA$Fixed.Data$TExPosition.Data$t[2],3), 
                            "%", ", p=",resBADA$Inference.Data$components$p.vals[2] ), 
                     limits = c(resBADA$Fixed.Data$Plotting.Data$constraints$miny, resBADA$Fixed.Data$Plotting.Data$constraints$maxy))+
  
  # Drawing axis lines 
  geom_line(aes(x=c(resBADA$Fixed.Data$Plotting.Data$constraints$minx, resBADA$Fixed.Data$Plotting.Data$constraints$maxx), 
                y=c(0,0)))+
  
  geom_line(aes(x=c(0,0), 
                y=c(resBADA$Fixed.Data$Plotting.Data$constraints$miny, resBADA$Fixed.Data$Plotting.Data$constraints$maxy)))
```

From tha above plot we can see that Africa and Europe is very much seperated from all other groups.  To find the variables that account for these differences, we examine the loadings of the variables(figure \@ref(fig:BADAVarMap)).

Factor map of individuals with tolerance intervals is shown in figure \@ref(fig:BADAIndMap). 

```{r BADAIndMap, out.width="75%", fig.align="center", fig.pos="H", warning=FALSE, message=FALSE, fig.cap="BADA Factor Scores (Individuals) with tolerance intervals", echo=FALSE, fig.height=5.8, fig.width=8}
FactorScore <- as.data.frame(resBADA$Fixed.Data$TExPosition.Data$fii[,c(1,2)])
FactorScore$labels <- labels(resBADA$Fixed.Data$TExPosition.Data$fii[,2])

p <- ggplot()

p +
  # Plotting the points 
  geom_point(aes(x=FactorScore$V1,
                   y=FactorScore$V2, 
                   color = designMatrix$Region)) +
  
  # Drawing the confidence ellipse 
  stat_ellipse(aes(x=FactorScore$V1,
                   y=FactorScore$V2, 
                   color = designMatrix$Region), 
               linetype = 6, type = "norm")+
  
  # Scaling the axis based on the plotting constraints 
  scale_x_continuous(name = paste0("Component 1 Inertia: ",
                                            round(resBADA$Fixed.Data$TExPosition.Data$t[1],3), "%", 
                                   ", p=", resBADA$Inference.Data$components$p.vals[1]) ) +
  scale_color_discrete(name= "Regions")+
  scale_y_continuous(name = paste0("Component 2 Inertia: ", 
                            round(resBADA$Fixed.Data$TExPosition.Data$t[2],3), 
                            "%", ", p=",resBADA$Inference.Data$components$p.vals[2] ))+
  
  # Drawing the axis 
  geom_line(aes(x=c(resBADA$Fixed.Data$Plotting.Data$constraints$minx, resBADA$Fixed.Data$Plotting.Data$constraints$maxx), 
                y=c(0,0)))+
  geom_line(aes(x=c(0,0),
                y=c(resBADA$Fixed.Data$Plotting.Data$constraints$miny, resBADA$Fixed.Data$Plotting.Data$constraints$maxy)))+
  
  # theme() is used for tweaking the final appearance of the plot 
   theme(legend.position = "bottom")
```

## Loadings

Figure \@ref(fig:BADAVarMap) shows the loadings of the BADA on World health risk. We can see that the loadings plot of BADA looks very much similar to that of the PCA Loadings plot. 

```{r BADAVarMap, out.width="75%", fig.align="center", fig.pos="H", warning=FALSE, message=FALSE, fig.cap="BADA Loadings", echo=FALSE, fig.height=5, fig.width=8}
Loadings <- as.data.frame(resBADA$Fixed.Data$TExPosition.Data$fj[,c(1,2)])
Loadings$labels <- labels(resBADA$Fixed.Data$TExPosition.Data$fj[,2])
p <- ggplot()

p + 
  # Plotting the points 
  geom_point(aes(x=Loadings$V1,
                   y=Loadings$V2)) +
  
  # Printing the labels 
  geom_text_repel(aes(x=Loadings$V1,
                      y=Loadings$V2, 
                      label = Loadings$labels)) +
  
  # Scaling the axis
  scale_x_continuous(name = paste0("Component 1 Inertia: ",
                                            round(resBADA$Fixed.Data$TExPosition.Data$t[1],3), "%", 
                                   ", p=", resBADA$Inference.Data$components$p.vals[1]), 
                     limits = c(resBADA$Fixed.Data$Plotting.Data$constraints$minx, resBADA$Fixed.Data$Plotting.Data$constraints$maxx) ) +
  
  scale_color_discrete(name= "Regions")+
  
  scale_y_continuous(name = paste0("Component 2 Inertia: ", 
                            round(resBADA$Fixed.Data$TExPosition.Data$t[2],3), 
                            "%", ", p=",resBADA$Inference.Data$components$p.vals[2] ), 
                     limits = c(resBADA$Fixed.Data$Plotting.Data$constraints$miny, resBADA$Fixed.Data$Plotting.Data$constraints$maxy))+
  
  # Drawing the axis lines 
  geom_line(aes(x=c(resBADA$Fixed.Data$Plotting.Data$constraints$minx, resBADA$Fixed.Data$Plotting.Data$constraints$maxx), 
                y=c(0,0)))+
  geom_line(aes(x=c(0,0), 
                y=c(resBADA$Fixed.Data$Plotting.Data$constraints$miny, resBADA$Fixed.Data$Plotting.Data$constraints$maxy)))
```

## Confusion Matrix

A confusion matrix helps us to understand the quality of the model in making predictions. The confusion matrix (Fixed Model) for BADA is shown in figure \@ref(fig:BADAConf).

```{r BADAConf, echo=FALSE, fig.cap="Confusion Matrix (Fixed Model)", fig.pos="H", fig.align="center"}
require(ggpubr)
ggtexttable(as.data.frame(resBADA$Fixed.Data$TExPosition.Data$assign$confusion))
```

## Reliability and Stability of the Analysis

Figure \@ref(fig:BADAConfGrp) shows the factor map of BADA for groups with 95% bootstrap confidence intervals.

```{r BADAConfGrp, out.width="75%", fig.align="center", fig.pos="H", warning=FALSE, message=FALSE, fig.cap="BADA Factor Scores (Group) with confidence intervals", echo=FALSE, fig.height=5.8, fig.width=8}
Fi <- as.data.frame(resBADA$Fixed.Data$TExPosition.Data$fi[,c(1,2)])
Fi$labels <- labels(resBADA$Fixed.Data$TExPosition.Data$fi[,2])
# Converting the variables into dataframe 
FactorScore <- as.data.frame(resBADA$Fixed.Data$TExPosition.Data$fii[,c(1,2)])
FactorScore$labels <- labels(resBADA$Fixed.Data$TExPosition.Data$fii[,2])
p <- ggplot()

p + 
  # Plotting the points 
  geom_point(aes(x=Fi$V1,
                   y=Fi$V2, color=unique(designMatrix$Region))) +
  
  # Drawing confidence ellipse 
  stat_conf_ellipse(aes(x=FactorScore$V1,
                   y=FactorScore$V2, 
                   color = designMatrix$Region), 
               linetype = 6, inherit.aes = F, geom = "path")+
  
  # Scalling axis 
  scale_x_continuous(name = paste0("Component 1 Inertia: ",
                                            round(resBADA$Fixed.Data$TExPosition.Data$t[1],3), "%", 
                                   ", p=", resBADA$Inference.Data$components$p.vals[1]),
                     limits = c(resBADA$Fixed.Data$Plotting.Data$constraints$minx, resBADA$Fixed.Data$Plotting.Data$constraints$maxx) ) +
  scale_color_discrete(name= "Regions")+
  scale_y_continuous(name = paste0("Component 2 Inertia: ", 
                            round(resBADA$Fixed.Data$TExPosition.Data$t[2],3), 
                            "%", ", p=",resBADA$Inference.Data$components$p.vals[2] ), 
                     limits = c(resBADA$Fixed.Data$Plotting.Data$constraints$miny, resBADA$Fixed.Data$Plotting.Data$constraints$maxy))+
  
  # Drawing axis lines 
  geom_line(aes(x=c(resBADA$Fixed.Data$Plotting.Data$constraints$minx, resBADA$Fixed.Data$Plotting.Data$constraints$maxx), 
                y=c(0,0)))+
   geom_line(aes(x=c(0,0),
                y=c(resBADA$Fixed.Data$Plotting.Data$constraints$miny, resBADA$Fixed.Data$Plotting.Data$constraints$maxy)))+
  
  # theme() is used for tweaking the final appearance of the plot 
   theme(legend.position = "bottom")
```

### Bootstrap Ratios 

```{r echo=FALSE, fig.width=6, fig.height=3, warning=FALSE, message=FALSE, fig.align="center", fig.pos="H", fig.cap="BADA Bootstrap Ratios"}


bootRatios <- as.data.frame(resBADA$Inference.Data$boot.data$fj.boot.data$tests$boot.ratios)
bootRatios$labels <- row.names(resBADA$Fixed.Data$TExPosition.Data$fj)
# Bootstrap for component 1
B1<- ggbarplot(bootRatios, 
          x= "labels", y="V1", 
          sort.val = "asc", 
          palette = "jco", 
          fill = "dodgerblue3", color = "white", 
          ggtheme = theme_minimal(), 
          title = "Component 1", 
          ylab = "Boot Ratios", xlab = "Variables", rotate=T)+ 
  
  geom_hline(yintercept = 2, linetype=2, color="brown1") +  
  geom_hline(yintercept = -2, linetype=2, color="brown1")


bootColors2 <- c("dodgerblue3","dodgerblue3",
                "dodgerblue3","dodgerblue3",
                "gray", "dodgerblue3")

# Bootstrap for component 2
B2<- ggbarplot(bootRatios, x= "labels", y="V2",
          sort.val = "asc", palette = "jco", 
          fill = bootColors2, color = "white", 
          ggtheme = theme_minimal(), 
          title = "Component 2", 
          ylab = "Boot Ratios", xlab = "", 
          rotate=T)+
  geom_hline(yintercept = 2, linetype=2, color="brown1") + 
  geom_hline(yintercept = -2, linetype=2, color="brown1")

multiplot(B1,B2, cols = 2)
```

From the above figure we can see that all the variables were significant for component one whereas only BMI_Male was insignificant for component two.

### Confusion Matrix (Random Model)

A jackknife procedure was used inorder to evaluate the generalization capacity of the analysis. This gave the following random effect confusion matrix: 

```{r BADAConfRand, echo=FALSE, fig.cap="Confusion Matrix (Random Model)", fig.pos="H", fig.align="center", fig.height=5, fig.width=12}
ggtexttable(as.data.frame(resBADA$Inference.Data$loo.data$loo.confuse))
```


`r if (knitr:::is_html_output()) '# References {-}'`
