# (PART) Multi Table Analysis {-}

# DiSTATIS

```{r include=FALSE}
# A clean start
rm(list=ls())
graphics.off()

# Loading required libraries 
library(DistatisR)
library(ggplot2)
library(PTCA4CATA)

# Loading the datasets 
Musical_Sorting <- read.csv("Data/natural36_constrained_forR(4).csv", row.names = 1)
Sorting_Data <- Musical_Sorting[-38,]
Design_Data <- read.csv("Data/sortingsubjectdesign.csv", row.names = 1)
Sorted  <- (Musical_Sorting[c(1:36),])
musicalExp <- Sorting_Data[]
Design <- read.csv("Data/design_musical.csv", row.names = 1)

# DistanceFromSort() converts the data matrix into a matrix of distances 
DistanceCube <- DistanceFromSort(Sorted)

# distatis() is a function from the package DistatisR that do the distance scaling
resDistasis <- distatis(DistanceCube)

# Bootstrap Sampling
Boots <- BootFactorScores(resDistasis$res4Splus$PartialF)

# Permutation test
permDistatis <- perm4PTCA(DistanceCube)
```

DISTATIS is a generalization of classical multidimensional scaling. Its goal is to analyze several distance matrices computed on the same set of objects. DISTATIS first evaluates the similarity between distance matrices. From this analysis, a compromise matrix is computed which represents the best aggrigate of the original matrices. The original distance matrices are then projected onto the compromise[@DiSTATIS].

To illustrate DISTATIS we will use the `Musical Sorting` Data. Three different "composers" are compared, each of them computing a distance matrix between the subjects. Heatmap of the distance cube for the given dataset is shown in figure \@ref(fig:DISTATISHeat)

```{r DISTATISHeat, fig.width=7, fig.height=7, out.width="50%", fig.align="center", fig.cap="Heatmap of S Plus Matrix", fig.pos="H", echo=FALSE, message=FALSE, warning=FALSE}
# RColorBrewer helps to choose colors from a palatte
require(RColorBrewer)

# Rowv,Colv = NA prevents the function from clustering the variables 
corrplot::corrplot(resDistasis$res4Cmat$C, method = "color", cl.pos = "n", tl.col = "black")
        #col= colorRampPalette(brewer.pal(5, "Blues"))(256))
```


## Computation

The given dataset is first convereted into an indicator matrix $L$ and this indicator matrix is transformed into a co-occurance matrix as: $$\mathbf{R_{[t]} = L_{[t]}L_{[t]}^T}$$ The co-occurance matrix is transformed then into a distance matrix by sustracting the co-occurance matrix from a conformable matrix filled with 1's. $$\mathbf{D_{[t]}=1-R_{[t]}}$$

Distance matrices cannot be analyzed directly and need to be transformed. This step corresponds to MDS and transforms a distance matrix into a corss-product matrix.

The corss product matrix denoted by $\mathbf{\hat{S}}$
 is obtained as $$\mathbf{\hat{S}=-\frac{1}{2} \Xi D \Xi^T}$$ where $\mathbf{\Xi = 1- 1m^T}$, $\mathbf{m}$ is the vector of mass. 
 
The compromise matrix is a cross product matrix that gives the best compromise of the studies. It is obtained as a weighted average of the study cross-product matrices. The weights are chosen so that studies agreeing the most with other studies will have the larger weights. 

## Comparing the studies

To analyze the similarity structure of the studies we start by creating a between study cosine matrix $\mathbf{C}$. This cosine, aslo known as the $\mathbf{R_V}$ coefficient is defined as $$\mathbf{R_V} = \frac{trace\{ \mathbf{S_{[t]}^T S_{[t']} }\}}{\sqrt{trace\{ \mathbf{S_{[t]}^T S_{[t]} }\} \times trace\{ \mathbf{S_{[t']}^T S_{[t']} }\}}}$$

## PCA of the cosine matrix 

The cosine matrix has the following eigendecomposition: $$\mathbf{C=P \Theta P^T}$$ with $\mathbf{P^TP = I}$, where $\mathbf{P}$ is the matrix of eigenvectors and $\mathbf{\Theta}$ is the diagonal matrix of the eigenvalues of $\mathbf{C}$. The Scree Plot is shown in figure \@ref(fig:DISTATISScree). 


```{r plotScreeDISTATIS, include=FALSE}

# Create a function to plot the scree
# ev: the eigen values to plot. no default
# max.ev the max eigen value
#        needed because ExPosition does not return all ev
#        but only the requested one. but return all tau
#        so if max.ev is specified, it is used to recompute
#        all eigenvalues
# p.ep: the probabilities associated to the ev
# alpha: threshold for significance. Default = .05
# col.ns  = color for ns ev. Default is Green
# col.sig = color for significant ev. Default is Violet
PlotScree <- function(ev,p.ev=NULL,max.vp=NULL,
                      alpha=.05,
                      col.ns = '#006D2C',col.sig='#54278F',
                      title = "Explained Variance per Dimension"
){
  # percentage of inertia
  val.tau = (100*ev/sum(ev))
  Top.y = ceiling(max(val.tau)*.1)*10
  # if ev is already a percentage convert it back
  if (!is.null(max.vp)){ev = ev*(max.vp/ev[1])}
  p <- ggplot(ev, aes(x = c(1:35), y= ev))
  p + 
    
    # drawing a line that passes through all the ploted points
    geom_line() + 
    
    # Printing y-axis label
    ylab("Inertia Extracted by the Components") +
    scale_x_continuous(name='Dimensions', breaks = c(1,2,3,4,5,6,7,8,9,10,11,12,13,14,15,16,17,18,19,20,21,22,23,24,25,26,27,28,29,30,31,32,33,34,35))  +
    geom_line() + 
    
    scale_y_continuous(sec.axis = sec_axis(~./sum(ev)*100,name="Percentage of Explained Variance")) +
    
    geom_point(col=ifelse(p.ev<alpha, 'blue', 'indianred3'), size=2) +theme_get()
   
} # end of function PlotScree
``` 
 
 
 `r if (knitr:::is_html_output()) '# References {-}'`
 
```{r DISTATISScree, out.width="60%", fig.align="center", fig.cap="Scree Plot for DISTATIS", fig.pos='H', fig.width=7.5, fig.height=4.5, echo=FALSE}

PlotScree(as.data.frame(permDistatis$fixedEigenvalues), p.ev = permDistatis$pEigenvalues)
```

```{r include=FALSE}
require(ggplot2)
require(ggpubr)

# A function for plotting the Distatis Rv
ggGraphDistatisRv = function(RvFS, axis1=1, axis2=2, ZeTitle="Distatis-Rv Map",
                             participant.colors = NULL, nude=F, RvCtr = NULL, lab = NULL)
  { 
  # if there is no colors passes, assigns colors based on design matrix 
  if (is.null(participant.colors)){
    participant.design <- diag(dim(RvFS)[1])
    participant.colors <- as.matrix(createColorVectorsByDesign(participant.design)$oc)
  }
  LeMinimum = apply(RvFS, 2, min)
  LeMaximum = apply(RvFS, 2, max)
  petitx = min(c(0, LeMinimum[axis1]))
  petity = LeMinimum[axis2]
  grandx = LeMaximum[axis1]  
  grandy = LeMaximum[axis2]
  fudgeFact_H = (grandx - petitx)/9
  fudgeFact_V = (grandy - petity)/9
  constraints <- list(minx = petitx - fudgeFact_H, maxx = grandx + 
        fudgeFact_H, miny = petity - fudgeFact_V, maxy = grandy + 
        fudgeFact_V)
    if (is.null(RvCtr)) {
        RvF2 <- RvFS^2
        RvSF2 <- apply(RvF2, 2, sum)
        RvCtr <- t(t(RvF2)/RvSF2)
    }
    p <- ggplot()
    
    # Plotting the points 
    p + geom_point(aes(x=RvFS[axis1], y=RvFS[axis2]), 
                    size=3, label=row.names(resDistasis$res4Cmat$G)) +
      geom_text_repel(aes(x=RvFS[axis1], y=RvFS[axis2]), label=row.names(resDistasis$res4Cmat$G))+
      
      # Printing thr title 
      ggtitle(ZeTitle) + 
      
      # Scaling the x and y axis
      scale_x_continuous(name = "", limits = c(minx = petitx - fudgeFact_H,
                                               maxx = grandx + 
        fudgeFact_H)) + 
      
      scale_y_continuous(name="", limits = c(miny = petity - fudgeFact_V, 
                                             maxy = grandy + 
        fudgeFact_V))+ 
      
     # Drawing the x and y axis 
       geom_vline(xintercept = 0)+
      geom_hline(yintercept = 0)
}
```

An element of a given eigenvector represents the projection of one study on this eigenvector. Thus the $T$ studies can be represented as points in the eigenspace and their similarities analyzed visually. Figure \@ref(fig:DISTATISRvMap) displays the projection of subjects onto the first and second component.

```{r DISTATISRvMap, fig.height=5, fig.width=9, echo=FALSE, warning=FALSE, message=FALSE, fig.pos="H", fig.align="center", fig.cap="Plot of the between-composers space", out.width="75%"}
ggGraphDistatisRv(as.data.frame(resDistasis$res4Cmat$G)) 
```


```{r include=FALSE}

# A function for plotting the compromise 
ggGraphDistatisCompromise = function(FS, axis1 = 1, axis2 = 2, constraints = NULL, item.colors = NULL,ZeTitle = "Distatis-Compromise", nude = FALSE, Ctr = NULL, lab=NULL)
  { # Assigning colors with respect to design matrix if no color details passed 
  if (is.null(item.colors)) {
        item.design <- diag(dim(FS)[1])
        item.colors <- as.matrix(createColorVectorsByDesign(item.design)$oc)
  }
  
  # Computing the constraints 
    if (is.null(constraints) || sum(names(constraints) %in% c("minx", 
        "maxx", "miny", "maxy")) != 4) {
        print("Making constraints")
        real.minimum <- min(FS)
        real.maximum <- max(FS)
        real.value <- max(c(abs(real.minimum), abs(real.maximum)))
        constraints <- list(minx = -real.value, maxx = real.value, 
            miny = -real.value, maxy = real.value)
    }
    if (is.null(Ctr)) {
        F2 <- FS^2
        SF2 <- apply(F2, 2, sum)
        Ctr <- t(t(F2)/SF2)
    }
  p <- ggplot()
  p + 
    
    # Plotting the points 
    geom_point(aes(x=FS[axis1], y=FS[axis2],colour = item.colors ), 
                 size=3)+
    
    # Printing the title 
    ggtitle(ZeTitle)+
      
    # Scaling the axis based on the plotting constraints 
    scale_x_continuous(name = "", limits = c(minx = -real.value,
                                               maxx = real.value)) + 
      scale_y_continuous(name="", limits = c(miny = -real.value, 
                                             maxy = real.value)) +
    
    # Drawing the x and y axis 
    geom_vline(xintercept = 0)+
    geom_hline(yintercept = 0)+
    
    # Printing labels
    geom_text_repel(aes(x=FS[axis1], y=FS[axis2], 
                        label=lab), size=3)
    
}
```

## Computing the compromise 

As for STATIS the weights are obtained by dividing each element of $\mathbf{p_1}$ by their sum. The vector containing these weights is denoted $\mathbf{\alpha}$. With $\alpha_t$ denoting the weight for the t-th study, the compromise matrix, denoted $$\mathbf{S}_{[+]} = \sum_t^T \alpha_t \mathbf{S}_{[t]}$$

The eigendecomposition of the compromise is: $$\mathbf{S}_{[+]} = \mathbf{V \Lambda V^T}$$ and the compromise factor scores for the observations are computed as: $$\mathbf{F} =\mathbf{V \Lambda^{\frac{1}{2} } }$$ 

The compromise plot of the music is shown in figure \@ref(fig:DISTATISCompromise). The music are color coded based on their composer.

```{r DISTATISCompromise, fig.height=8, fig.width=9, echo=FALSE, warning=FALSE, fig.pos="H", fig.align="center", fig.cap="Analysis of the compromise: Plot of the subjects in the plane defined by the first two components of the compromise matrix", out.width="75%", message=FALSE}
labelsCompromise <- as.data.frame(labels(resDistasis$res4Splus$F)[[1]])
require(ggrepel)

ggGraphDistatisCompromise(FS = as.data.frame(resDistasis$res4Splus$F) ,
                          lab = labelsCompromise, ZeTitle = "" , item.colors = Design$composer)
```


## Bootstrap 

A bootstrap sampling was conducted to compute the confidence intervals and the compromise plot with the bootstrap intervals is shown in figure \@ref(fig:DISTATISCompromiseBoot)

```{r DISTATISCompromiseBoot, fig.height=8, fig.width=9, echo=FALSE, warning=FALSE, fig.pos="H", fig.align="center", fig.cap="Compromise with Bootstraped Confidence Intervals", out.width="75%", message=FALSE}
# ggGraphDistatisCompromise() is a user defined function for plotting the Compromise 
ggGraphDistatisCompromise(FS = as.data.frame(resDistasis$res4Splus$F) ,
                          lab = labelsCompromise, ZeTitle = "", 
                          item.colors = Design$composer) + 
  # stat_conf_ellipse draws the confidence ellipse 
  stat_conf_ellipse(aes(x=as.data.frame(resDistasis$res4Splus$F)[,1],
                        y=as.data.frame(resDistasis$res4Splus$F)[,2], 
                        color=Design$composer), 
                    inherit.aes = T, bary = T, linetype=2)
```

