# MFA/STATIS

Multiple Factor Analysis (MFA) analyzes observations described by several blocks or sets of variables. MFA is performed in two steps. First a principal component analysis is performed on each data set which is then normalized by dividing all its elements by the square root of the first eigenvalue obtained from its PCA. Second, the normalized dataset are merged to form a unique matrix and a global PCA is performed on this matrix. The individual data sets are then projected onto the global analysis to analyze communalities and discrepancies[@MFA]. 

To illustrate MFA, we use the `World Health` dataset. The heatmap of the dataset is shown in figure \@ref(fig:WorldCorPlot)

## Computation 

For our exaple we have four sub-tables. Let's call them $\mathbf{X}_{[1]}$, $\mathbf{X}_{[2]}$, $\mathbf{X}_{[3]}$ and $\mathbf{X}_{[4]}$. The first singular value is the normalizing factor used to divide the elements of the data table. For example, the PCA of the first group gives a first eigenvalue $\varphi_1$. This gives the first normalized data matrix $\mathbf{Z}_{[1]}$: $$\mathbf{Z}_{[1]} = \varphi_1^{-1} \times \mathbf{X}_{[1]}$$. Similarly matrices $\mathbf{Z}_{[2]}$, $\mathbf{Z}_{[3]}$ and $\mathbf{Z}_{[4]}$ are computed. The Scree plot of our given dataset is shown in figure \@ref(fig:MFAScree)


```{r include=FALSE, warning=FALSE, message=FALSE}
# Clean Start
rm(list = ls()) 
graphics.off() 

library(TInPosition)
require(ggplot2)

# Loading Dataset
WorldHealth <- read.csv("Data/WorldHealth.csv", row.names = 1)

# Cleaning Data 
names(WorldHealth) <- c("Fires", "Drownings", "Poisonings", "Falls",
                        "T.Chole_F", "T.Chole_M",
                        "BP_F", "BP_M",
                        "BMI_F","BMI_M",
                        "GenGovExp", "TotalExp_GDP", 
                        "PerCapitaExp_XchangeRate","PerCapita_PPP",
                        "IMM_MCV", "IMM_DTP3")

# A function for plotting multiple plots 
multiplot <- function(..., plotlist=NULL, file, cols=1, layout=NULL) {
  library(grid)
  
  # Make a list from the ... arguments and plotlist
  plots <- c(list(...), plotlist)
  
  numPlots = length(plots)
  
  # If layout is NULL, then use 'cols' to determine layout
  if (is.null(layout)) {
    # Make the panel
    # ncol: Number of columns of plots
    # nrow: Number of rows needed, calculated from # of cols
    layout <- matrix(seq(1, cols * ceiling(numPlots/cols)),
                     ncol = cols, nrow = ceiling(numPlots/cols))
  }
  
  if (numPlots==1) {
    print(plots[[1]])
    
  } else {
    # Set up the page
    grid.newpage()
    pushViewport(viewport(layout = grid.layout(nrow(layout), ncol(layout))))
    
    # Make each plot, in the correct location
    for (i in 1:numPlots) {
      # Get the i,j matrix positions of the regions that contain this subplot
      matchidx <- as.data.frame(which(layout == i, arr.ind = TRUE))
      
      print(plots[[i]], vp = viewport(layout.pos.row = matchidx$row,
                                      layout.pos.col = matchidx$col))
    }
  }
}

Healthtable <- read.csv("Data/worldhealthtabledesign.csv")
Healthtable2 <- read.csv("Data/worldhealthtabledesign2.csv")
designMatrix <- read.csv("Data/design.csv")
library(MExPosition)
require(factoextra)
require(ggplot2)
designforpartial <- c(rep("RiskFactors", 175), rep("Spendings", 175), rep("Immunization", 175))

designregion <- read.csv("Data/designpartial.csv")
WorldHealth <- WorldHealth[,c(5:16)]
Healthtable2 <- Healthtable2[,c(5:16)]
Healthtable <- Healthtable[c(5:16),]
```

```{r include=FALSE, warning=F, message=FALSE}
resMFA <- mpMFA(WorldHealth, column.design = Healthtable2, DESIGN = designMatrix$Region, make.design.nominal = T, graphs = F, make.columndesign.nominal = T)
```

```{r plotScreeMFA, include=FALSE}

# Create a function to plot the scree
# ev: the eigen values to plot. no default
# max.ev the max eigen value
#        needed because ExPosition does not return all ev
#        but only the requested one. but return all tau
#        so if max.ev is specified, it is used to recompute
#        all eigenvalues
# p.ep: the probabilities associated to the ev
# alpha: threshold for significance. Default = .05
# col.ns  = color for ns ev. Default is Green
# col.sig = color for significant ev. Default is Violet
PlotScree <- function(ev,p.ev=NULL,max.vp=NULL,
                      alpha=.05,
                      col.ns = '#006D2C',col.sig='#54278F',
                      title = "Explained Variance per Dimension"
){
  # percentage of inertia
  val.tau = (100*ev/sum(ev))
  Top.y = ceiling(max(val.tau)*.1)*10
  # if ev is already a percentage convert it back
  if (!is.null(max.vp)){ev = ev*(max.vp/ev[1])}
  p <- ggplot(ev, aes(x = c(1:12), y= ev))
  p + 
    geom_line() + ylab("Inertia Extracted by the Components") +
    scale_x_continuous(name='Dimensions', breaks = c(1,2,3,4,5,6,7,8,9,10,11,12))  +
    geom_line() + 
    scale_y_continuous(sec.axis = sec_axis(~./sum(ev)*100,name="Percentage of Explained Variance")) +
    geom_point( size=2) +theme_get() 
  
} # end of function PlotScree

```

```{r MFAScree, out.width="60%", fig.align="center", fig.cap="Scree Plot for MFA", fig.pos='H', fig.width=7.5, fig.height=4.5, echo=FALSE}
PlotScree(as.data.frame(resMFA$mexPosition.Data$Table$eigs))
```

These normalized matrices are concatenated into an $I \times T$ matrix called the global data matrix dentoed as $\mathbf{Z}$. $$\mathbf{Z} = [ \mathbf{Z_{[1]}} \ \ \mathbf{Z_{[2]}} \ \ \mathbf{Z}_{[3]} \ \ \mathbf{Z}_{[4]} ]$$

## Computing the global PCA

To analyze the global matrix, we use standard PCA. This amounts to computing the singular value decomposition of the global data matrix: $$\mathbf{Z=U \Delta V^T}$$ and the global factor scores are obtained as: $$\mathbf{F = M^{-\frac{1}{2}}U \Delta} $$ 

The factoe scores of the given data table is shown in figure \@ref(fig:MFAFMap1) (Components 1 and 2) and \@ref(fig:MFAFMap2) (Components 3 and 4)

```{r MFAFMap1,  out.width="70%", fig.align="center", fig.pos="H", warning=FALSE, message=FALSE, fig.cap="MFA World Health Characteristics. Factor scores of the observations plotted on the first two components", echo=FALSE}
require(tibble)

# Converting table into data frame for passing it to ggplot
FactorScore <- as.data.frame(resMFA$mexPosition.Data$Table$fi)[c(1,2)]
FactorScore$labels <- labels(resMFA$mexPosition.Data$Table$fi[,1])
cbbPalette <- c("#000000", "#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "darkslategray1", "blueviolet")
# Functions to plot Factor Scores 
p <- ggplot()      # Plots the base
png(filename = "MFAFactor.png", width = 1000, height = 800, units = "px")
p +
  
  # geom_point () plots the data points
  geom_point(aes(x=FactorScore[,1],y=FactorScore[,2], 
                 color=designMatrix$Region), 
             position = "dodge", size=4)  + geom_text_repel(aes(x=FactorScore[,1],y=FactorScore[,2]), label=FactorScore$labels)+
  
  # scale_x_continuous() scales the plotting constraints of x axis
  scale_x_continuous(limits = c(-0.3,0.3) ,name = paste0("Component 1 Inertia: ",
                                                         round(resMFA$mexPosition.Data$Table$t[1],3), "%")) + 
  
  
  # scale_y_continuous() scales the plotting constraints of y axis
  scale_y_continuous(limits = c(-0.2,0.2),name = paste0("Component 2 Inertia: ", 
                                                        round(resMFA$mexPosition.Data$Table$t[2],3), "%")) + 
  
  # geom_hline() prints a horizontal line at the specified y intercept
  geom_hline(yintercept = 0)+
  
  # geom_vline() prints a vertical line at the specified y intercept
  geom_vline(xintercept = 0)+
  
  # theme() is used for tweaking the final appearance of the plot 
  theme(legend.position = "bottom")+ scale_colour_manual(values=cbbPalette, name="Regions")+ theme_minimal()+
  
  # theme() is used for tweaking the final appearance of the plot 
  theme(legend.position = "bottom")  + theme(text = element_text(size=25)) + ggtitle("Factor Scores") 
dev.off()
```

From the above figure we can see that component 1 is seperating Africa from Europe and component 2 is seperating Asia from Europe

```{r MFAFMap2,  out.width="70%", fig.align="center", fig.pos="H", warning=FALSE, message=FALSE, fig.cap="MFA World Health Characteristics. Factor scores of the observations plotted on components three and four", echo=FALSE}
require(tibble)

# Converting table into data frame for passing it to ggplot
FactorScore <- as.data.frame(resMFA$mexPosition.Data$Table$fi)[c(3,4)]
FactorScore$labels <- labels(resMFA$mexPosition.Data$Table$fi[,1])

# Functions to plot Factor Scores 
p <- ggplot()      # Plots the base
p +
  
  # geom_point () plots the data points
  geom_point(aes(x=FactorScore[,1],y=FactorScore[,2], 
                 color=designMatrix$Region), 
             position = "dodge")  +  
  
  # scale_x_continuous() scales the plotting constraints of x axis
  scale_x_continuous(name = paste0("Component 3 Inertia: ",
                                   round(resMFA$mexPosition.Data$Table$t[3],3), "%")) + 
  scale_color_discrete(name= "Regions")+
  
  # scale_y_continuous() scales the plotting constraints of y axis
  scale_y_continuous(name = paste0("Component 4 Inertia: ", 
                                   round(resMFA$mexPosition.Data$Table$t[4],3), "%")) + 
  
  # geom_hline() prints a horizontal line at the specified y intercept
  geom_hline(yintercept = 0)+
  
  # geom_vline() prints a vertical line at the specified y intercept
  geom_vline(xintercept = 0)+
  
  # theme() is used for tweaking the final appearance of the plot 
  theme(legend.position = "bottom")
```

## Partial Factor Scores 

```{r MFAPartialFMap1,  out.width="70%", fig.align="center", fig.pos="H", warning=FALSE, message=FALSE, fig.cap="MFA World Health Characteristics. Partial Factor scores of the observations plotted on the first two components", echo=FALSE}
require(tibble)

# Converting table into data frame for passing it to ggplot
FactorScore <- as.data.frame(resMFA$mexPosition.Data$Table$partial.fi)[c(1,2)]
FactorScore$labels <- labels(resMFA$mexPosition.Data$Table$partial.fi[,1])

# Functions to plot Factor Scores 
p <- ggplot()      # Plots the base
p +
  
  # geom_point () plots the data points
  geom_point(aes(x=FactorScore[,1],y=FactorScore[,2], color=designregion$Region, shape=designforpartial))  +  
  
  # scale_x_continuous() scales the plotting constraints of x axis
  scale_x_continuous(name = paste0("Component 1 Inertia: ",
                                   round(resMFA$mexPosition.Data$Table$t[1],3), "%")) + 
  scale_color_discrete(name= "Regions")+
  
  # scale_y_continuous() scales the plotting constraints of y axis
  scale_y_continuous(name = paste0("Component 2 Inertia: ", 
                                   round(resMFA$mexPosition.Data$Table$t[2],3), "%")) + 
  
  # geom_hline() prints a horizontal line at the specified y intercept
  geom_hline(yintercept = 0)+
  
  # geom_vline() prints a vertical line at the specified y intercept
  geom_vline(xintercept = 0)+
  
  # theme() is used for tweaking the final appearance of the plot 
  theme(legend.position = "bottom")
```

```{r MFAPartialFMap2,  out.width="70%", fig.align="center", fig.pos="H", warning=FALSE, message=FALSE, fig.cap="MFA World Health Characteristics. Partial Factor scores of the observations plotted on components three and four", echo=FALSE}
require(tibble)

# Converting table into data frame for passing it to ggplot
FactorScore <- as.data.frame(resMFA$mexPosition.Data$Table$partial.fi)[c(3,4)]
FactorScore$labels <- labels(resMFA$mexPosition.Data$Table$partial.fi[,1])

# Functions to plot Factor Scores 
p <- ggplot()      # Plots the base
p +
  
  # geom_point () plots the data points
  geom_point(aes(x=FactorScore[,1],y=FactorScore[,2], color=designregion$Region, shape=designforpartial))  +  
  
  # scale_x_continuous() scales the plotting constraints of x axis
  scale_x_continuous(name = paste0("Component 3 Inertia: ",
                                   round(resMFA$mexPosition.Data$Table$t[3],3), "%")) + 
  scale_color_discrete(name= "Regions")+
  
  # scale_y_continuous() scales the plotting constraints of y axis
  scale_y_continuous(name = paste0("Component 4 Inertia: ", 
                                   round(resMFA$mexPosition.Data$Table$t[4],3), "%")) + 
  
  # geom_hline() prints a horizontal line at the specified y intercept
  geom_hline(yintercept = 0)+
  
  # geom_vline() prints a vertical line at the specified y intercept
  geom_vline(xintercept = 0)+
  
  # theme() is used for tweaking the final appearance of the plot 
  theme(legend.position = "bottom")
```

## Loadings 

```{r MFALoading,  out.width="70%", fig.align="center", fig.pos="H", warning=FALSE, message=FALSE, fig.cap="MFA World Health Characteristics. Loadings of the observations plotted on the first two components", echo=FALSE}
require(tibble)
require(ggrepel)
# Converting table into data frame for passing it to ggplot
FactorScore <- as.data.frame(resMFA$mexPosition.Data$Table$Q)[c(1,2)]
FactorScore$labels <- labels(resMFA$mexPosition.Data$Table$Q[,1])

# Functions to plot Factor Scores 
p <- ggplot()      # Plots the base
p +
  
  # geom_point () plots the data points
  geom_point(aes(x=FactorScore[,1],y=FactorScore[,2], color=Healthtable$Grp), 
             position = "dodge") + 
  geom_text_repel(data = FactorScore, 
                  mapping = aes(x=FactorScore[,1],y=FactorScore[,2], label = FactorScore$labels))  +  
  scale_color_discrete(name="Sub Tables")+
  # scale_x_continuous() scales the plotting constraints of x axis
  scale_x_continuous(name = paste0("Component 1 Inertia: ",
                                   round(resMFA$mexPosition.Data$Table$t[1],3), "%"))+
  
  # scale_y_continuous() scales the plotting constraints of y axis
  scale_y_continuous(name = paste0("Component 2 Inertia: ", 
                                   round(resMFA$mexPosition.Data$Table$t[2],3), "%")) + 
  
  # geom_hline() prints a horizontal line at the specified y intercept
  geom_hline(yintercept = 0)+
  
  # geom_vline() prints a vertical line at the specified y intercept
  geom_vline(xintercept = 0)+
  
  # theme() is used for tweaking the final appearance of the plot 
  theme(legend.position = "bottom")
```

From the above figure we can see that variables within each sub-table is very well correlated as they are very near to each other in the Loadings map. 



`r if (knitr:::is_html_output()) '# References {-}'`
